The effects of multi-dimensional drought on land cover change and vegetation productivity in continental Chile

### 2024-03-19 article/paper_drought_Chile2023.tex
update article

\usepackage{lineno}\linenumbers \usepackage{multirow} \usepackage{lscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The effects of multidimensional drought on land cover change and vegetation productivity in continental Chile},
  pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Meza; Iongel Duran-Llacer; Francisco Fernández; Alejandro Venegas-González; Nicolas Raab; Dylan Craven},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{The effects of multidimensional drought on land cover change and
vegetation productivity in continental Chile}
\author[1]{Francisco Zambrano%
\corref{cor1}%
\fnref{fn1}}
 \ead{francisco.zambrano@umayor.cl} 
\author[2]{Anton Vrieling%
%
}
 \ead{a.vrieling@utwente.nl} 
\author[]{Francisco Meza%
%
}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, enhanced]}{\end{tcolorbox}}\fi

@@

From 1960 to 2019, land use change has impacted around one-third of the
Earth's surface, which is four times more than previously thought
\citep{Winkler2021}. Multiple studies aim to analyze and forecast
changes in land cover globally \citep{Winkler2021, Song2018} and
regionally \citep{Chamling2020, Homer2020, Yang2021}. Some others seek
to analyze the impact of land cover change on climate conditions such as
temperature and precipitation \citep{Luyssaert2014, Pitman2012}. There
is less research on the interaction between drought and land cover
change \citep{Chen2022, Akinyemi2021, Peng2017}. \citet{Peng2017}
conducted a worldwide investigation utilizing net primary production to
examine the spatial and temporal variations in vegetation productivity
at global level. The study aimed to assess the influence of drought by
comparing the twelve-month Standardized Precipitation Evapotranspiration
Index (SPEI) and land cover change. According to their findings, drought
is responsible for 37\% of the decline in vegetation productivity, while
water availability accounts for 55\% of the variation. \citet{Chen2022}
studied the trend of vegetation greenness and productivity and its
relation to meteorological drought (SPEI of twelve months in December)
and soil moisture at the global level. The results showed lower
correlations (\textless0.2) for both variables. \citet{Akinyemi2021}
evaluates drought trends and land cover change using vegetation indices
in Botwsana in a semi-arid climate. These studies mostly looked at how
changes in land cover and vegetation productivity are related to a
single drought index (SPEI) over a single time period of 12 months. SPEI
takes into account the combined effect of precipitation and AED as a
water balance, but it does not allow us to know the contribution of each
variable on its own. Some things worth investigating in terms of land
cover change and vegetation productivity are: i) How do they respond to
short- to long-term meteorological droughts? ii) How do they behave in
humid and arid climatic zones regarding drought? And iii) What is the
role of soil moisture? Likewise, there is a lack of understanding of how
the alteration in water supply and demand is affecting land cover
transformations.

Chile's diverse climatic and ecosystem types
(\citet{Beck2023};\citet{Luebert2022}) make it an ideal natural
laboratory for studying climate and ecosystems. Additionally, the
country has experienced severe drought conditions that have had
significant effects on vegetation and water storage. Central Chile faced
a persistent precipitation deficit between 2010 and 2022, defined as a
megadrought \citep{Garreaud2017}, which has impacted the Chilean
ecosystem. This megadrought was defined by the Standardized
Precipitation Index (SPI) of twelve months in December having values
below one standard deviation. Some studies have addressed how this
drought affects single ecosystems in terms of forest development
\citep{Miranda2020, Venegas2018}, forest fire occurrence
\citep{UrrutiaJalabert2018}, and crop productivity
\citep{Zambrano2023, Zambrano2018, Zambrano2016}. We found one study
regarding land cover and drought in Chile. The study by
\citet{Fuentes2021} evaluates water scarcity and land cover change in
Chile between 29° and 39° of south latitude. \citet{Fuentes2021} used
the SPEI of one month for evaluating drought, which led to misleading
results. For example, they did not find a temporal trend in the SPEI but
found a decreasing trend in water availability and an increase trend on
AED, which in turn should have been capable of being captured with
longer time scales of the SPEI. The term ``megadrought'' in Chile is
used to describe a prolonged water shortage that lasts for several
years, resulting in a permanent deficit that impacts the hydrological
system \citep{Boisier2018}. Hence, it is imperative to assess temporal
scales that take into account the cumnulative effect within some years.
There is little knowledge about the relationship between drought and
ecosystem in Chile; thus, it is important to understand in more detail
how meteorological and soil moisture droughts influence ecosystem
dynamics to inform adaptation options.

A detailed spatiotemporal assessment of the interaction of drought for
short- to long-term and land cover change requires information on
vegetation as well as weather variables such as precipitation,
temperature, and soil moisture. Weather networks in Chile present some
disadvantages, such as spatio-temporal gaps, a short history, and
irregular quality, which make them difficult to represent the whole
extent of the country spatially. In order to do this, we use reanalysis
data from ERA5-Land \citep{MunozSabater2021} to create drought indices
that consider AED, precipitation, and soil moisture over a range of time
periods, from the short to the long term. Also, we use vegetation
spectral information and annual land cover change from the
Moderate-Resolution Imaging Spectroradiometer (MODIS). We expect to gain
insight regarding the temporal evolution of water demand, water supply,
and soil moisture, as well as the interaction with land cover change and
vegetation productivity. Here, we analyze the multi-dimensional impacts
of drought across ecosystems in continental Chile. More specifically, we
aim to assess: i) temporal changes in land-use cover and the direction
and magnitude of their relationships with drought indices for water
demand and supply, soil moisture, and vegetation productivity; ii)
short- to long-term temporal trends in multi-scalar drought indices; and
iii) the relationship between vegetation productivity and drought
indices for water demand and supply and soil moisture across Chilean
ecosystems.

@@

Continental Chile has a diverse climate conditions with strong gradients
from north to south and east to west \citep{Aceituno2021}
(Figure~\ref{fig-studyArea} a), which determines its great ecosystem
diversity (\citet{Luebert2022}) (Figure~\ref{fig-studyArea} c). The
Andes Mountains are a main factor in climate latitudinal variation
\citep{Garreaud2009}. In order to characterize the climate and ecosystem
of Chile, we utilize the Köppen-Geiger classification system developed
by \citet{Beck2023} and the land cover data derived from the MODIS
product for the period of 2001--2022, based on the International
Geosphere-Biosphere Programme (IGBP) classification scheme proposed by
\citet{Friedl2019}. ``Norte Grande'' and ``Norte Chico'' predominate in
an arid desert climate with hot (Bwh) and cold (Bwk) temperatures. At
the south of ``Norte Chico,'' the climate changes to an arid steppe with
cold temperatures (Bsk). In these two northern regions, the land is
mostly bare, with a minor surface of vegetation types such as shrubland
and grassland. In the zones ``Centro'' and the north half of ``Sur,''
the main climate is Mediterranean, with warm to hot summers (Csa and
Csb). Land cover in ``Centro'' comprises a significant amount of
shrubland and savanna (50\%), grassland (16\%), forest (8\%), and
croplands (5\%). An oceanic climate (Cfb) predominates in the south of
``Sur'' and the north of ``Austral.'' Those zones are high in forest and
grassland. The southern part of the country has a tundra climate, and in
``Austral'', it is a cold semi-arid area with an extended surface of
grassland, forest, and, to a lesser extent, savanna.

@@

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection
6.1 from MODIS. This product has a yearly frequency from 2001 to 2022.
The IGBP defines 17 classes; from these, we regrouped into ten
macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9
to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands,
13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
Thus, we have a land cover raster time series with the ten classes for
2001 and 2023. Prior to using this class, we compared it with the types
of land cover made by \citet{Zhao2016}, which is a more detailed land
cover map of Chile with a 30 m spatial resolution for 2012--2013. It
reached a global accuracy of \textasciitilde0.82. The procedure of
validation is demonstrated in Section S1 of the supplementary
material.To derive a proxy for vegetation productivity, we used the
MOD13A3 collection 6.1 product from MODIS \citep{Didan2015}. It provides
vegetation indices (NDVI and EVI) at 1km of spatial resolution and
monthly frequency. The MOD13A3 and MCD12Q1 were retrieved from the
online Data Pool, courtesy of the NASA EOSDIS Land Processes Distributed
Active Archive Center (LP DAAC), USGS Earth Resources Observation and
Science (EROS) Center, Sioux Falls, South Dakota,
https://lpdaa.usgs.gov/tools/data-pool/.

\begin{table}[!ht]
\caption{Description of the earth observation data used }
\label{tab-desEOD}
\small
\centering
\begin{tabular}{p{0.13\textwidth}cp{0.3\textwidth}p{0.095\textwidth}ccc}
\hline
\multirow{1}{*}{\centering Product} & Sub-product & Variable & Spatial Resolution  & Period & Units & Short Name \\ 
\hline
\multirow{4}{*}{ERA5L} & ~ & Precipitation & \multirow{4}{*}{~0.1°} & \multirow{4}{*}{1981-2023} & mm & P \\ 
         &  & Maximum temperature & ~ & & $°C$ & $T_{max}$ \\ 
         &  & Minimum temperature & ~ & & $°C$ & $T_{min}$ \\ 
         &  & Volumetric Soil Water Content at 1m & ~ & & $m3/m3$ & SM \\ 
ERA5L* & & Atmospheric Evaporative Demand & 0.1° & 1981-2023 & mm & AED \\
        \multirow{2}{*}{MODIS} & MOD13A3.061 & Normalized Difference Vegetation Index & \multirow{2}{*}{~1 km} & 2000-2023 & ~ & NDVI \\ 
         & MCD12Q1.061 & land cover IGBP scheme & & 2001-2022 & ~ & land cover \\ 
\hline
\end{tabular}
{\raggedright *Derived from ERA5L with Eq. \ref{eq-AED}. \par}
\end{table}

For soil moisture, water supply, and water demand variables, we used
ERA5L \citep{MunozSabater2021}, a reanalysis dataset that provides the
evolution of the meteorological and soil moisture variables since 1950.
It has a spatial resolution of 0.1° (9 km), hourly frequency, and global
coverage. We selected the variables for total precipitation, maximum and
minimum temperature at 2 meters, and volumetric soil water layers
between 0 and 100cm of depth (layer 1 to layer 3).

\hypertarget{land-cover-change-and-trend}{%
\subsection{Land cover change and
trend}\label{land-cover-change-and-trend}}

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection
6.1 from MODIS. \citet{Zambrano2018} and \citet{Fuentes2021} have
previously used this product for studies of drought and land cover. The
MCD12Q1 has a yearly frequency from 2001 to 2022. The IGBP defines 17
classes; from these, we regrouped into ten macroclasses, as follows:
classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as
grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as
snow and ice, 16 as barren, and 17 to water bodies. Thus, we have a land
cover raster time series with the ten macroclasses for 2001 and 2023. We
validate the land cover macroclasses regarding a highly detailed (30 m
of spatial resolution) land cover map made for Chile by \citet{Zhao2016}
for 2013-2014. Our results showed a global accuracy of
\textasciitilde0.82 and a F1 score of \textasciitilde0.66. Section S2 in
the Supplementary Material shows the procedure for validation.

\hypertarget{trend-of-drought-and-interaction-to-land-cover-change}{%
\subsection{Trend of drought and interaction to land cover
change}\label{trend-of-drought-and-interaction-to-land-cover-change}}

@@

In order to compute the drought indices that uses water demand, it is
necessary to first calculate the AED. To do this, we employed the
Hargreaves technique \citep{Hargreaves1994, Hargreaves1985}, by applying
the following equation:

@@

where \(Ra\) \((MJ\,m^2\, day^{-1})\) is extraterrestrial radiation;
\(T\), \(T_{max}\), and \(T_{min}\) are mean, maximum, and minimum
temperature \((°C)\). We calculate the centroid coordinates per pixel
and use the latitude to estimate \(Ra\).

We chose the method of Hargreaves to estimate AED because of its
simplicity, which only requires temperatures and extrarrestrial
radiation. Also, it has been recommended over other methods (e.g.,
Penman-Monteith) when the access to climatic variables is limited
\citep{Vicente-Serrano2014}.

@@

To derive the drought indices of water supply and demand, soil moisture,
and vegetation we used the ERA5L dataset and the MODIS product, with a
monthly frequency for 1981-2023 and 2000-2023, respectively.

@@

For the purpose of monitoring water supply drought, we used the
well-known Standardized Precipitation Index (SPI), which the World
Meteorological Organization (WMO) recommended. The SPI solely relies on
precipitation data. Also, it has been used worldwide for the study of
drought, including in Chile (\citet{Garreaud2017};
\citet{Zambrano2017}). The primary cause of drought is precipitation
anomalies, and temperature usually makes it worse (Luo et al.~2017).
Nowadays, there is an increase in attention toward using water demand
separately to monitor droughts. (\citet{Vicente-Serrano2020};
\citet{Noguera2022}). Thus, to evaluate water demand, we chose the
Evaporative Demand Drought Index (EDDI), developed by Hobbins et
al.~(016) and McEvoy et al.~(2016), which is based on the AED. EDDI is
currently used for monitoring drought in the United States
(https://psl.noaa.gov/eddi/). In our case, we used only temperature for
AED, a difference from the original formulation of EDDI, which also
considered wind besides temperature. To consider the combined effect of
water supply and demand, we selected the SPEI, which corresponds to a
balance between precipitation and AED. \citet{Vicente-Serrano2010}
proposed the SPEI, and it has improved the SPI by incorporating
temperature for drought monitoring. For SPEI, an auxiliary variable D =
P-AED is calculated. Soil moisture is the main driver of vegetation
productivity, particularly in semi-arid regions (\citet{Li2022}). Hence,
for soil water drought, we used the SSI (Standardized Soil Moisture
Index) (Hao and AghaKouchak 2013; A. AghaKouchak 2014) which is a
multi-scale index similar to SPI, SPEI, and EDDI. In our case, for the
SSI, we used the average soil moisture from ERA5L at 1m depth. Finally,
for the proxy of productivity, we used the zcNDVI proposed by Zambrano
et al.~(2018) which will be derived from the NDVI retrieved from
MOD13A1.

@@

\hypertarget{intercation-drought-indices-and-land-cover-change}{%
\subsubsection{Intercation drought indices and land cover
change}\label{intercation-drought-indices-and-land-cover-change}}

@@

@@

\hypertarget{land-cover-change-and-trend-1}{%
\section{Land cover change and
trend}\label{land-cover-change-and-trend-1}}

\hypertarget{trend-of-drought-and-interaction-to-land-cover-change-1}{%
\subsection{Trend of drought and interaction to land cover
change}\label{trend-of-drought-and-interaction-to-land-cover-change-1}}

\hypertarget{trend-of-drought-indices-1}{%
\subsubsection{Trend of drought
indices}\label{trend-of-drought-indices-1}}

@@

>>>>

\usepackage{lineno}\linenumbers \usepackage{multirow} \usepackage{lscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The effects of multi-dimensional drought on land cover change and vegetation productivity in continental Chile},
  pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Meza; Iongel Duran-Llacer; Francisco Fernández; Alejandro Venegas-González; Nicolas Raab; Dylan Craven},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{The effects of multi-dimensional drought on land cover change and
vegetation productivity in continental Chile}
\author[1]{Francisco Zambrano%
\corref{cor1}%
\fnref{fn1}}
 \ead{francisco.zambrano@umayor.cl} 
\author[2]{Anton Vrieling%
%
}
 \ead{a.vrieling@utwente.nl} 
\author[]{Francisco Meza%
%
}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, sharp corners, frame hidden, enhanced, interior hidden, breakable]}{\end{tcolorbox}}\fi

@@

From 1960 to 2019, land use change has impacted around one-third of the
Earth's surface, which is four times more than previously thought
\citep{Winkler2021}. Multiple studies aim to analyze and forecast
changes in land cover globally \citep{Winkler2021, Song2018} and
regionally \citep{Chamling2020, Homer2020, Yang2021}. Some others seek
to analyze the impact of land cover change on climate conditions such as
temperature and precipitation \citep{Luyssaert2014, Pitman2012}. There
is less research on the interaction between drought and land cover
change \citep{Chen2022, Akinyemi2021, Peng2017}. \citet{Peng2017}
conducted a worldwide investigation utilizing net primary production to
examine the spatial and temporal variations in vegetation productivity
at global level. The study aimed to assess the influence of drought by
comparing the twelve-month Standardized Precipitation Evapotranspiration
Index (SPEI) and land cover change. According to their findings, drought
is responsible for 37\% of the decline in vegetation productivity, while
water availability accounts for 55\% of the variation. \citet{Chen2022}
studied the trend of vegetation greenness and productivity and its
relation to meteorological drought (SPEI of twelve months in December)
and soil moisture at the global level. The results showed lower
correlations (\textless0.2) for both variables. \citet{Akinyemi2021}
evaluates drought trends and land cover change using vegetation indices
in Botwsana in a semi-arid climate. These studies mostly looked at how
changes in land cover and vegetation productivity are related to a
single drought index (SPEI) over a single time period of 12 months. SPEI
takes into account the combined effect of precipitation and AED as a
water balance, but it does not allow us to know the contribution of each
variable on its own. Some things worth investigating in terms of land
cover change and vegetation productivity are: i) How do they respond to
short- to long-term meteorological and soil moisture droughts? ii) How
is the drought impacting land cover changes? And iii) How do they behave
in humid and arid climatic zones regarding drought? Likewise, there is a
lack of understanding of how the alteration in water supply and demand
is affecting land cover transformations.

For monitoring drought, the World Meteorological Organization recommends
the SPI (Standardized Precipitation Index) \citep{WMO2012}. The SPI is a
multi-scalar drought index that only uses precipitation to assess short-
to long-term droughts. The primary cause of drought is precipitation
anomalies, and temperature usually makes it worse \citep{Luo2017}.
Nowadays, there is an increase in attention toward using AED separately
to monitor droughts \citep{Vicente-Serrano2020}. One reason is due to
its attribution to increasing flash droughts in water-limited regions
\citep{Noguera2022}. \citet{Vicente-Serrano2010} proposed the
Standardized Precipitation Evapotranspiration Index (SPEI), which
incorporated the temperature effect by subtracting AED from
precipitation. SPEI allows for analysis of the combined effect of
precipitation and AED. Since its formulation, it has been used worldwide
for the study and monitoring of drought
\citep{Gebrechorkos2023, Liu2021}. \citet{Hobbins2016} and
\citet{McEvoy2016} developed the Evaporative Demand Drought Index (EDDI)
to monitor droughts solely using the AED, and it has proven effective in
monitoring flash droughts \citep{Li2024, Ford2023}. For soil moisture,
several drought indices exist, such as the Soil Moisture Deficit Index
(SDMI) \citep{Narasimhan2005} and the Soil Moisture Agricultural Drought
Index (SMADI) \citep{Souza2021}. \citet{Hao2013} and
\citet{AghaKouchak2014} proposed the Standardized Soil Moisture Index
(SSI), which has a similar formulation as the SPI, SPEI, and EDDI. Thus,
there are plenty of drought indices that allow for a comprehensive
assessment of drought from short- to long-term scales and consider
variables from the earth's water balance on their own (e.g.,
precipitation, AED, soil moisture). Using this information, we can
advance our understanding of the impact of drought on ecosystems.

Chile's diverse climatic and ecosystem types
\citep{Beck2023, Luebert2022} make it an ideal natural laboratory for
studying climate and ecosystems. Additionally, the country has
experienced severe drought conditions that have had significant effects
on vegetation and water storage. North-central Chile has faced a
persistent precipitation deficit since 2010, defined as a mega drought.
\citep{Garreaud2017}, which has impacted the Chilean ecosystem. This
megadrought was defined by the Standardized Precipitation Index (SPI) of
twelve months in December having values below one standard deviation.
Some studies have addressed how this drought affects single ecosystems
in terms of forest development \citep{Miranda2020, Venegas2018}, forest
fire occurrence \citep{UrrutiaJalabert2018}, and crop productivity
\citep{Zambrano2023, Zambrano2018, Zambrano2016}. We found one study
regarding land cover and drought in Chile. The study by
\citet{Fuentes2021} evaluates water scarcity and land cover change in
Chile between 29° and 39° of south latitude. \citet{Fuentes2021} used
the SPEI of one month for evaluating drought, which led to misleading
results. For example, they did not find a temporal trend in the SPEI but
found a decreasing trend in water availability and an increase trend on
AED, which in turn should have been capable of being captured with
longer time scales of the SPEI. The term ``megadrought'' in Chile is
used to describe a prolonged water shortage that lasts for several
years, resulting in a permanent deficit that impacts the hydrological
system \citep{Boisier2018}. Therefore, it is crucial to evaluate
temporal scales that consider the cumulative impact over a period of
several years. There is limited understanding of the correlation between
drought and the environment in Chile. Therefore, it is crucial to have a
deeper understanding of how meteorological and soil moisture droughts
impact the dynamics of the environment in order to make informed
decisions on how to adapt.

Here, we analyze the multi-dimensional impacts of drought across
ecosystems in continental Chile. More specifically, we aim to assess: i)
short- to long-term temporal trends in multi-scalar drought indices; ii)
temporal changes in land-use cover and the direction and magnitude of
their relationships with trends in drought indices; and iii) the trend
in vegetation productivity and its relationship with drought indices
across Chilean ecosystems.

@@

Continental Chile has diverse climate conditions with strong gradients
from north to south and east to west \citep{Aceituno2021}
(Figure~\ref{fig-studyArea} a), which determines its great ecosystem
diversity \citep{Luebert2022} (Figure~\ref{fig-studyArea} c). The Andes
Mountains are a main factor in climate latitudinal variation
\citep{Garreaud2009}. ``Norte Grande'' and ``Norte Chico'' predominate
in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures. At
the south of ``Norte Chico,'' the climate changes to an arid steppe with
cold temperatures (Bsk). In these two northern regions, the land is
mostly bare, with a small surface of vegetation types such as shrubland
and grassland. In the zones ``Centro'' and the north half of ``Sur,''
the main climate is Mediterranean, with warm to hot summers (Csa and
Csb). Land cover in ``Centro'' comprises a significant amount of
shrubland and savanna (50\%), grassland (16\%), forest (8\%), and
croplands (5\%). An oceanic climate (Cfb) predominates in the south of
``Sur'' and the north of ``Austral.'' Those zones are high in forest and
grassland. The southern part of the country has a tundra climate, and in
``Austral,'' it is a cold semi-arid area with an extended surface of
grassland, forest, and, to a lesser extent, savanna.

@@

To analyze land cover change, we use the classification scheme by the
IGBP (International Geosphere-Biosphere Programme) from the product
MCD12Q1 collection 6.1 from MODIS. To derive a proxy for vegetation
productivity, we used the Normalized Difference Vegetation Index (NDVI)
from the product MOD13A3 collection 6.1 from MODIS \citep{Didan2015}.
MOD13A3 provides vegetation indices at 1km of spatial resolution and
monthly frequency. The NASA EOSDIS Land Processes Distributed Active
Archive Center (LP DAAC), USGS Earth Resources Observation and Science
(EROS) Center, Sioux Falls, South Dakota, provided the MOD13A3 and
MCD12Q1 from the online Data Pool, accessible at
https://lpdaa.usgs.gov/tools/data-pool/.

\begin{table}[!ht]
\caption{Description of the satellite and reanalysis data used}
\label{tab-desEOD}
\small
\centering
\begin{tabular}{p{0.13\textwidth}cp{0.3\textwidth}p{0.095\textwidth}ccc}
\hline
\multirow{1}{*}{\centering Product} & Sub-product & Variable & Spatial Resolution  & Period & Units & Short Name \\ 
\hline
\multirow{4}{*}{ERA5L} & ~ & Precipitation & \multirow{4}{*}{~0.1°} & \multirow{4}{*}{1981-2023} & mm & P \\ 
         &  & Maximum temperature & ~ & & $°C$ & $T_{max}$ \\ 
         &  & Minimum temperature & ~ & & $°C$ & $T_{min}$ \\ 
         &  & Volumetric Soil Water Content at 1m & ~ & & $m3/m3$ & SM \\ 
ERA5L* & & Atmospheric Evaporative Demand & 0.1° & 1981-2023 & mm & AED \\
        \multirow{2}{*}{MODIS} & MOD13A3.061 & Normalized Difference Vegetation Index & \multirow{2}{*}{~1 km} & 2000-2023 & ~ & NDVI \\ 
         & MCD12Q1.061 & land cover IGBP scheme & & 2001-2022 & ~ & land cover \\ 
\hline
\end{tabular}
{\raggedright *Calculated from maximum and minimum temperatures derived from ERA5L with Eq. \ref{eq-AED}. \par}
\end{table}

For soil moisture, water supply, and water demand variables, we used
ERA5L (ECMWF Reanalysis version 5 over land) \citep{MunozSabater2021}, a
reanalysis dataset that provides the evolution of atmospheric and land
variables since 1950. It has a spatial resolution of 0.1° (9 km), hourly
frequency, and global coverage. We selected the variables for total
precipitation, maximum and minimum temperature at 2 meters, and
volumetric soil water layers between 0 and 100cm of depth (layer 1 to
layer 3). Table \ref{tab-desEOD} shows a summary of the data and its
main characteristics.

\hypertarget{trend-of-short--to-long-term-drought}{%
\subsection{Trend of short- to long-term
drought}\label{trend-of-short--to-long-term-drought}}

@@

In order to compute the drought indices that use water demand, it is
necessary to first calculate the AED. To do this, we employed the
Hargreaves method \citep{Hargreaves1994, Hargreaves1985} by applying the
following equation:

@@

where \(Ra\) \((MJ\,m^2\, day^{-1})\) is extraterrestrial radiation;
\(T\), \(T_{max}\), and \(T_{min}\) are mean, maximum, and minimum
temperature \((°C)\) at 2m. For calculating \(Ra\) we used the
coordinate of the latitud of the centroid of each pixel. We chose the
method of Hargreaves to estimate AED because of its simplicity, which
only requires temperatures and extrarrestrial radiation. Also, it has
been recommended over other methods (e.g., Penman-Monteith) when the
access to climatic variables is limited \citep{Vicente-Serrano2014}.

@@

To derive the drought indices of water supply and demand, soil moisture,
and vegetation (i.e., the proxy of productivity), we used the ERA5L
dataset and the MODIS product, with a monthly frequency for 1981--2023
and 2000--2023, respectively.

@@

For the purpose of monitoring water supply drought, we used the
well-known Standardized Precipitation Index (SPI), which the World
Meteorological Organization (WMO) recommended. The SPI solely relies on
precipitation data. Also, it has been used worldwide for the study of
drought, including in Chile (\citet{Garreaud2017};
\citet{Zambrano2017}). The primary cause of drought is precipitation
anomalies, and temperature usually makes it worse \citep{Luo2017}.
Nowadays, there is an increase in attention toward using water demand
separately to monitor droughts. (\citet{Vicente-Serrano2020};
\citet{Noguera2022}). Thus, to evaluate water demand, we chose the
Evaporative Demand Drought Index (EDDI), developed by
\citet{Hobbins2016} and \citet{McEvoy2016}, which is based on the AED.
EDDI is currently used for monitoring drought in the United States
(https://psl.noaa.gov/eddi/). In our case, we used only temperature for
AED, a difference from the original formulation of EDDI, which also
considered wind besides temperature. To consider the combined effect of
water supply and demand, we selected the SPEI, which corresponds to a
balance between precipitation and AED. \citet{Vicente-Serrano2010}
proposed the SPEI, and it has improved the SPI by incorporating
temperature for drought monitoring. For SPEI, an auxiliary variable D =
P-AED is calculated. Soil moisture is the main driver of vegetation
productivity, particularly in semi-arid regions (\citet{Li2022}). Hence,
for soil water drought, we used the SSI (Standardized Soil Moisture
Index) \citep{Hao2013, AghaKouchak2014} which is a multi-scale index
similar to SPI, SPEI, and EDDI. In our case, for the SSI, we used the
average soil moisture from ERA5L at 1m depth. Finally, for the proxy of
productivity, we used the zcNDVI proposed by \citet{Zambrano2018} which
will be derived from the NDVI retrieved from MOD13A1.

@@

\hypertarget{interaction-of-land-cover-and-drought}{%
\subsection{Interaction of land cover and
drought}\label{interaction-of-land-cover-and-drought}}

\hypertarget{land-cover-change}{%
\subsubsection{Land cover change}\label{land-cover-change}}

To analyze the land cover change, we use the IGBP scheme from the
MCD12Q1 collection 6.1 from MODIS. \citet{Zambrano2018} and
\citet{Fuentes2021} have previously used this product for studies of
drought and land cover in Chile. The MCD12Q1 has a yearly frequency from
2001 to 2022. The IGBP defines 17 classes; from these, we regrouped into
ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands,
8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to
croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to
water bodies. Thus, we have a land cover raster time series with the ten
macroclasses for 2001 and 2023. We validate the land cover macroclasses
regarding a highly detailed (30 m of spatial resolution) land cover map
made for Chile by \citet{Zhao2016} for 2013-2014. Our results showed a
global accuracy of \textasciitilde0.82 and a F1 score of
\textasciitilde0.66. Section S2 in the Supplementary Material shows the
procedure for validation.

\hypertarget{relationship-between-land-cover-and-drought-trends}{%
\subsubsection{Relationship between land cover and drought
trends}\label{relationship-between-land-cover-and-drought-trends}}

@@

\hypertarget{drought-impacts-on-vegetation-productivity}{%
\subsection{Drought impacts on vegetation
productivity}\label{drought-impacts-on-vegetation-productivity}}

@@

\hypertarget{trend-of-short--to-long-term-drought-1}{%
\subsection{Trend of short- to long-term
drought}\label{trend-of-short--to-long-term-drought-1}}

@@

\hypertarget{interaction-of-land-cover-and-drought-1}{%
\subsubsection{Interaction of land cover and
drought}\label{interaction-of-land-cover-and-drought-1}}

\hypertarget{land-cover-change-1}{%
\subsection{Land cover change}\label{land-cover-change-1}}
